---
title: "Upstream and downstream compartments"
author: "Ada Yan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: TRUE
    keep_tex: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE)
```

# Run whole document with k0 = TRUE and k0 = FALSE, for k = 0 and k > 0 cases

```{r}
k0 <- FALSE
```

```{r include = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
# load R package with model fitting functions
devtools::load_all("~/git_repos/cellkineticmodels/") # change this to the directory of the git repository
# working directory should be scripts/ relative to the directory of the git repository
# i.e. the folder this vignette is in

if(k0) {
  dir_name <- "source_term_model_k0_files/" # save output files and plots here
} else {
  dir_name <- "source_term_model_files/"
}
if(!dir.exists(dir_name)) {
  dir.create(dir_name)
}
options(mc.cores = parallel::detectCores()) # parallelisation of inference
rstan::rstan_options(auto_write = TRUE) # avoids recompilation of Stan files for inference
set.seed(1) # set seed for random number generation

# filenames for saving
samples_filename <- paste0(dir_name, "samples.rds")
pred_filename <- paste0(dir_name, "pred.rds")
sim_data_filename <- paste0(dir_name, "sim_data.rds")
p_filename <- paste0(dir_name, "p.rds")
```

# Figure 2

## Simulate data

```{r}
# stan model with which to simulate data: model with upstream and downstream compartments
pred_model <- "differentiation_model_fix_U_Lb_pred"

label_end <- 49.; # end of labelling period

f <- 0.032 # fraction of label in drinking water
delta <- 0.070 # turnover rate of body water
b_w <- 5.82 # normalisation for body water

pred_times <- c(3, seq(0, 98, by = 7)) %>% sort # times for which to generate data
sigma_L <- 0.005 # standard deviation of lognormal noise on data
```

# Kinetically heterogeneous upstream compartment

## Sample parameter values to simulate data

```{r}
# define transformed parameters and their ranges
raw_par_names <- c("log10_p_C", "log10_p_E", "log10_dstar_C_on_p_C_minus_Delta", 
                   "d_E_scaled", "dstar_E_scaled", "log10_Delta_on_p_C", "k")
par_ranges <- matrix(c(-4, -2,
                       -4, -2,
                       0, 2,
                       0, 1,
                       0, 1,
                       -2, 0,
                       0, 21),
                     nrow = 2)
colnames(par_ranges) <- raw_par_names
rownames(par_ranges) <- c("min", "max")
N_samples <- 100
par_names <- c("p_C", "p_E", "dstar_C", "dstar_E", "Delta", "two_k_minus_one", "C_on_E")
```

```{r eval = FALSE}
# LHS sampling of transformed parameters from uniform distribution defined by range above
raw_samples <- lhs::randomLHS(N_samples, length(raw_par_names))
colnames(raw_samples) <- raw_par_names
raw_samples[,c("d_E_scaled", "dstar_E_scaled")] <- 
  t(apply(raw_samples[,c("d_E_scaled", "dstar_E_scaled")], 1, sort))

scale_samples <- function(vec, bounds) {
  vec * diff(bounds) + bounds[1]
}

for(i in raw_par_names) {
  raw_samples[,i] <- scale_samples(raw_samples[,i], par_ranges[,i])
}

if(k0) {
  raw_samples[,"k"] <- 0
} else {
  raw_samples[,"k"] <- floor(raw_samples[,"k"])
}

samples <- raw_samples
colnames(samples) <- par_names

# untransform to retrieve original parameter values
samples[,"p_C"] <- 10^raw_samples[,"log10_p_C"]
samples[,"p_E"] <- 10^raw_samples[,"log10_p_E"]
samples[,"Delta"] <- 10^raw_samples[,"log10_Delta_on_p_C"] * samples[,"p_C"]
samples[,"dstar_C"] <- 10^raw_samples[,"log10_dstar_C_on_p_C_minus_Delta"] * (samples[,"p_C"] - samples[,"Delta"])
samples[,"dstar_E"] <- 10^(raw_samples[,"log10_p_E"] + (log10(delta) - raw_samples[,"log10_p_E"]) * raw_samples[,"dstar_E_scaled"])
samples[,"two_k_minus_one"] <- 2^raw_samples[,"k"] - 1
d_E <- 10^(raw_samples[,"log10_p_E"] + (log10(delta) - raw_samples[,"log10_p_E"]) * raw_samples[,"d_E_scaled"])
samples[,"C_on_E"] <- (d_E - samples[,"p_E"]) / samples[,"Delta"] / 2^raw_samples[,"k"]
saveRDS(samples, samples_filename)
```

## Solve for fraction of label in downstream compartment

```{r}
samples <- readRDS(samples_filename)
pred_times <- seq(0, 100)
data_list_pars <- lapply(par_names, function(x) array(samples[, x]))
names(data_list_pars) <- par_names

data_list <- c(data_list_pars,
               list(f = f,
                    delta = delta,
                    b_w = b_w,
                    label_end = label_end,
                    N_samples = N_samples,
                    T = length(pred_times),
                    ts = pred_times))
```

```{r eval = FALSE}
pred <- make_pred(pred_model, data_list) %>%
  extract_pred(N_samples = N_samples, pred_name = "Ehat") %>%
  as.data.frame %>% as_tibble %>%
  mutate(t = pred_times) %>%
  pivot_longer(-t, names_to = "run", values_to = "Ehat") %>%
  mutate(run = gsub("V", "", run),
         run = as.numeric(run))
saveRDS(pred, pred_filename)
```

## Add noise to generate simulated data

```{r eval = FALSE}
sim_data <- pred %>%
  mutate(Lhat = vnapply(Ehat, function(y) rlnorm(1, log(y), sigma_L)),
         Lhat = pmax(Lhat, 0))
saveRDS(sim_data, sim_data_filename)
```

## Fit one-compartment model to downstream data

```{r}
sim_data <- sim_data_filename %>% readRDS
fit_single <- function(run1, eval = TRUE) {
  fit_filename <- paste0(dir_name, "fit", run1, ".rds")
  if(file.exists(fit_filename) || !eval) {
    # if(!eval) {
    return(fit_filename)
  }
  sim_data <- sim_data %>% filter(run == run1 & t > 0)
  fit_data_list <- c(list(T = nrow(sim_data),
                          ts = sim_data$t,
                          L = sim_data$Lhat,
                          sigma = sigma_L,
                          frac = data_list$f),
                     data_list[c("label_end", "delta")])
  
  model <- "lymphocyte_model_fixed_U_lognormal_noise"
  
  fit <- fit_stan_defaults(model, fit_data_list)
  saveRDS(fit, fit_filename)
  invisible(fit)
}
```

```{r eval = FALSE}
invisible(lapply(seq_len(N_samples), fit_single, eval = TRUE))
```

## Compile file with sampled and fitted values

```{r}
samples <- readRDS(samples_filename) %>%
  as.data.frame %>% as_tibble %>%
  mutate(run = seq_len(nrow(samples)),
         turnover = (two_k_minus_one + 1) * Delta * C_on_E + p_E,
         production_rate_by_division = two_k_minus_one * Delta * C_on_E + p_E,
         influx_ratio = 1/(1 + p_E/Delta/(two_k_minus_one + 1) / C_on_E))
```

```{r eval = FALSE}
get_p_median <- function(run1) {
  fit <- run1 %>% fit_single(eval = FALSE) %>% readRDS
  fit <- as.matrix(fit$fit) %>% as.data.frame %>% as_tibble %>%
    select(pb_w, dstar) %>%
    mutate(run = run1,
           p = pb_w / b_w) %>%
    select(-pb_w)
  fit
}
p <- lapply(seq_len(N_samples), get_p_median) %>%
  do.call(rbind, .)
saveRDS(p, p_filename)
```

```{r eval = FALSE}
get_sum_stats <- function(run1) {
  fit <- run1 %>% fit_single(eval = FALSE) %>% readRDS
  fit <- as.matrix(fit$fit) %>% as.data.frame %>% as_tibble %>%
    select(pb_w, dstar) %>%
    mutate(p = pb_w / b_w) %>%
    select(-pb_w) %>%
    summarise_all(list(lower = function(x) quantile(x, probs = 0.025),
                       median = median,
                       upper = function(x) quantile(x, probs = 0.975))) %>%
    mutate(run = run1)
  fit
}
sum_stats <- lapply(seq_len(N_samples), get_sum_stats) %>%
  do.call(rbind, .) %>%
  full_join(samples) %>%
  rename(p_C = "p_C", p_E = "p_E", dstar_C = "dstar_C", dstar_E = "dstar_E",
         C_on_E = "C_on_E") %>%
  select(run, everything())
write.csv(sum_stats, paste0(dir_name, "source_term_model_estimated_pars.csv"), row.names = FALSE)
```

## Fig 3

```{r fig.width = 12, fig.height = 3}
p <- readRDS(p_filename) %>%
  group_by(run) %>%
  summarise(p_median = median(p), dstar_median = median(dstar)) %>%
  full_join(samples)

if(!k0) {
  p <- p %>% filter(two_k_minus_one > 0)
}

diag_tibble <- tibble(x = c(0, 1))

make_plot <- function(p, diag_tibble, x_var, y_var, x_lab, y_lab) {
  x_max <- p %>%
    select(c(x_var, y_var)) %>%
    as.matrix %>%
    max
  ggplot(p, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(aes(color = influx_ratio)) +
    geom_line(data = diag_tibble, aes(x = x, y = x)) +
    theme_bw() +
    coord_cartesian(xlim = c(0, x_max), ylim = c(0, x_max)) +
    scale_color_continuous(limits = c(0, 1)) +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          text = element_text(size = 14),
          legend.position = "none") +
    xlab(x_lab) +
    ylab(y_lab)
}

proliferation_plot <- make_plot(p, diag_tibble, "p_E", "p_median",
                                "True value of \n proliferation rate \n (per day)",
                                latex2exp::TeX("$\\hat{p}$ (per day)"))

turnover_plot <- make_plot(p, diag_tibble, "turnover", "p_median",
                           "True value of \n turnover rate \n (per day)",
                           latex2exp::TeX("$\\hat{p}$ (per day)"))

production_plot <- make_plot(p, diag_tibble, "production_rate_by_division", "p_median",
                             "True value of \n production rate by \n division (per day)",
                             latex2exp::TeX("$\\hat{p}$ (per day)"))

dstar_plot <- make_plot(p, diag_tibble, "dstar_E", "dstar_median",
                        "True value of \n disappearance  rate of \n labelled cells (per day)",
                        latex2exp::TeX("$\\hat{d^*}$ (per day)"))

g1 <- gridExtra::grid.arrange(proliferation_plot, turnover_plot, production_plot, dstar_plot, nrow = 1)
ggsave(paste0(dir_name, "fit_one_compartment_lhs.pdf"), g1, width = 12, height = 3)
g1
```

## Fig 4

```{r fig.width = 5, fig.height = 3}
dir_name_k0 <- "source_term_model_k0_files/" # save output files and plots here
dir_name <- "source_term_model_files/"

# filenames for saving
samples_filename <- paste0(dir_name, "samples.rds")
samples_k0_filename <- paste0(dir_name_k0, "samples.rds")
p_k0_filename <- paste0(dir_name_k0, "p.rds")
p_filename <- paste0(dir_name, "p.rds")

p <- readRDS(p_filename) %>%
  group_by(run) %>%
  summarise(p_median = median(p), dstar_median = median(dstar))
p_k0 <- readRDS(p_k0_filename) %>%
  group_by(run) %>%
  summarise(p_median = median(p), dstar_median = median(dstar))
p <- p %>%
  bind_rows(p_k0) %>%
  select(-run) %>%
  bind_cols(samples) %>%
    mutate(proliferation_error = abs(p_median - p_E)/p_E * 100,
         turnover_error = abs(p_median - turnover)/turnover * 100,
         prod_by_div_error = abs(p_median - production_rate_by_division)/production_rate_by_division * 100,
         k = log2(two_k_minus_one + 1),
         log2_k = log2(k))

theor_error <- expand_grid(influx_ratio = seq(0, 1, by = .01),
                           k = seq(0, 20)) %>%
  mutate(proliferation_error = influx_ratio / (1 - influx_ratio) * (2^k-1)/2^k * 100,
         turnover_error = influx_ratio/2^k * 100,
         prod_by_div_error = 0,
         log2_k = log2(k))

g <- ggplot(theor_error) +
  geom_line(aes(x = influx_ratio, y = proliferation_error, color = log2_k, group = k)) +
  scale_y_log10("Error in \n proliferation rate (%)", breaks = 10^seq(-2, 4), labels = scales::label_number()) +
  theme_bw() +
  geom_point(data = p, aes(x = influx_ratio, y = proliferation_error, color = log2_k, group = k), size = .5) +
  xlab("Influx ratio") +
  labs(color = "log2(k)")
ggsave(paste0(dir_name, "proliferation_error.pdf"), g, width = 3.5, height = 2)
g <- ggplot(theor_error, aes(x = influx_ratio, y = turnover_error, color = log2_k, group = k)) +
  geom_line() +
  theme_bw() +
  geom_point(data = p, aes(x = influx_ratio, y = turnover_error, color = log2_k, group = k), size = .5) +
  xlab("Influx ratio") +
  ylab("Error in \n turnover rate (%)") +
  labs(color = "log2(k)")
ggsave(paste0(dir_name, "turnover_error.pdf"), g, width = 3, height = 2)
g

g <- ggplot(theor_error, aes(x = influx_ratio, y = prod_by_div_error, color = log2_k, group = k)) +
  geom_line() +
  theme_bw() +
  geom_point(data = p, aes(x = influx_ratio, y = prod_by_div_error, color = log2_k, group = k), size = .5) +
  xlab("Influx ratio") +
  ylab("Error in production \n rate by division (%)") +
  labs(color = "log2(k)")
ggsave(paste0(dir_name, "prod_by_div_error.pdf"), g, width = 3, height = 2)
g
```

```{r}
```
This report is brought to you by `kirlabelling` commit `r system("git -C ~/git_repos/kirlabelling/ rev-parse --short HEAD", intern = TRUE)` and `kirlabbook` commit `r system("git -C ~/2020/kirlabbook/ rev-parse --short HEAD", intern = TRUE)`.