---
title: "Upstream and downstream compartments"
author: "Ada Yan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: TRUE
    keep_tex: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE)
```

# Run whole document with k0 = TRUE and k0 = FALSE, for k = 0 and k > 0 cases

```{r}
k0 <- FALSE
```

```{r include = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
# load R package with model fitting functions
devtools::load_all("~/git_repos/cellkineticmodels/") # change this to the directory of the git repository
# working directory should be scripts/ relative to the directory of the git repository
# i.e. the folder this vignette is in

if(k0) {
  dir_name <- "source_term_model_k0_files/" # save output files and plots here
} else {
  dir_name <- "source_term_model_files/"
}
if(!dir.exists(dir_name)) {
  dir.create(dir_name)
}
options(mc.cores = parallel::detectCores()) # parallelisation of inference
rstan::rstan_options(auto_write = TRUE) # avoids recompilation of Stan files for inference
set.seed(1) # set seed for random number generation

# filenames for saving
samples_filename <- paste0(dir_name, "samples.rds")
pred_filename <- paste0(dir_name, "pred.rds")
sim_data_filename <- paste0(dir_name, "sim_data.rds")
p_filename <- paste0(dir_name, "p.rds")
fit_homogeneous_filename <- paste0(dir_name, "fit_homogeneous.rds")
sim_data_homogeneous_filename <- paste0(dir_name, "sim_data_homogeneous.rds")
```

# Figure 2

## Simulate data

```{r}
# stan model with which to simulate data: model with upstream and downstream compartments
pred_model <- "differentiation_model_fix_U_Lb_pred"

label_end <- 49.; # end of labelling period

f <- 0.032 # fraction of label in drinking water
delta <- 0.070 # turnover rate of body water
b_w <- 5.82#4.41 # normalisation for body water

p_E <- 2.39e-3#2.89e-3
dstar_E <- 7.75e-3#2.10e-2
p_C <- 4.5e-4
dstar_C <- 3e-4
Delta <- p_C - dstar_C # kinetic homogeneity
k <- if(k0) 0 else 10 
d_E <- (dstar_E - p_E) * .9 + p_E # enforce p_E < d_E < dstar_E
C_on_E <- (d_E - p_E) / Delta / 2^k
pred_times <- c(3, seq(0, 98, by = 7)) %>% sort # times for which to generate data
sigma_L <- 0.005 # standard deviation of lognormal noise on data
data_list <- list(f = f,
                  delta = delta,
                  b_w = b_w,
                  p_C = array(p_C),
                  dstar_C = array(dstar_C),
                  p_E = array(p_E),
                  dstar_E = array(dstar_E),
                  Delta = array(Delta),
                  two_k_minus_one = array(2^k - 1),
                  C_on_E = array(C_on_E),
                  label_end = label_end,
                  T = length(pred_times),
                  ts = pred_times,
                  N_samples = 1)
```

```{r eval = TRUE}
pred <- make_pred(pred_model, data_list) # simulate fraction of label in upstream and downstream compartments

term_names <- paste0(c("Chat" ,"Ehat"))
terms <- lapply(term_names, extract_pred, pred = pred, N_samples = 1)
terms <- tibble(t = pred_times, Chat = as.numeric(terms[[1]]), Ehat = as.numeric(terms[[2]]))

# add lognormal noise
sim_data_homogeneous <- terms %>%
  mutate_at(c("Chat", "Ehat"), function(x) vnapply(x, function(y) rlnorm(1, log(y), sigma_L))) %>%
  mutate_at(c("Chat", "Ehat"), function(x) pmax(x, 0))
saveRDS(sim_data_homogeneous, sim_data_homogeneous_filename)
```

## Fit one-compartment model to data

```{r}
sim_data_homogeneous <- readRDS(sim_data_homogeneous_filename) %>%
  filter(t > 0)
fit_data_list <- c(list(T = nrow(sim_data_homogeneous),
                        ts = sim_data_homogeneous$t,
                        L = sim_data_homogeneous$Ehat,
                        sigma = sigma_L,
                        frac = data_list$f),
                   data_list[c("label_end", "delta")])

model <- "lymphocyte_model_fixed_U_lognormal_noise"
```

```{r eval = TRUE}
fit <- fit_stan_defaults(model, fit_data_list)
saveRDS(fit, fit_homogeneous_filename)
```

## Plot marginal posterior distributions (Fig 2)

```{r posterior_homogeneous, fig.width = 6, fig.height = 3, fig.cap = "Marginal posterior distributions for (left) $p_E$ and (right) $d^*_E$ when fitting the one-compartment model to the downstream data. The red line shows the values of (left) the mean turnover rate across all downstream cells and (right) the disappearance rate of labelled downstream cells used to simulate the data, while the blue line shows the proliferation rate of downstream cells used to simulate the data. \\label{fig:posterior_homogeneous}"}
fit <- readRDS(fit_homogeneous_filename)
fit_matrix <- as.matrix(fit$fit)
pred_single <- get_max_LL_row(fit$fit)
pred_single <- pred_single[grep("Lhat", names(pred_single))]
pred_tibble <- tibble(t = c(0, fit$data$ts),
                      Lhat = c(0, pred_single))
p <- fit_matrix[,"pb_w"] / b_w
theoretical_turnover <- Delta * 2^k * C_on_E + p_E
theoretical_production_rate_by_division <- Delta * (2^k - 1) * C_on_E + p_E
plot_tibble <- tibble(p = p, dstar = fit_matrix[,"dstar"])

palette1 <- ggpubfigs::friendly_pal("contrast_three")
light_grey <- "#808B96"

m <- ggplot(pred_tibble) +
  geom_line(aes(x = t, y = Lhat)) +
  geom_point(data = sim_data_homogeneous, aes(x = t, y = Ehat)) +
  theme_bw() +
  xlab("Time (days)") +
  ylab("Fraction of label") +
  labs(tag = ifelse(k0, "D", "A"))

g <- ggplot(plot_tibble, aes(x = p)) +
  geom_histogram(fill = light_grey) +
  geom_vline(xintercept = theoretical_production_rate_by_division, linetype = "solid", color = palette1[2]) +
  geom_vline(xintercept = theoretical_turnover, linetype = "dashed", color = palette1[1]) +
  geom_vline(xintercept = p_E, linetype = "dotdash", color = palette1[3]) +
  theme_bw() +
  xlab(latex2exp::TeX("$\\hat{p_E}")) +
  ylab("Frequency") +
  expand_limits(x = 0) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(tag = ifelse(k0, "E", "B"))

h <- ggplot(plot_tibble, aes(x = dstar)) +
  geom_histogram(fill = light_grey) +
  geom_vline(xintercept = dstar_E, linetype = "dashed", color = palette1[1]) +
  theme_bw() +
  xlab(latex2exp::TeX("$\\hat{d^*_E}$")) +
  ylab("Frequency") +
  expand_limits(x = 0) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(tag = ifelse(k0, "F", "C"))

g1 <- gridExtra::grid.arrange(m, g, h, nrow = 1)
ggsave(paste0(dir_name, "fit_homogeneous.pdf"), g1, width = 6, height = 2)
```

# Kinetically heterogeneous upstream compartment

## Sample parameter values to simulate data

```{r}
# define transformed parameters and their ranges
raw_par_names <- c("log10_p_C", "log10_p_E", "log10_dstar_C_on_p_C_minus_Delta", 
                   "d_E_scaled", "dstar_E_scaled", "log10_Delta_on_p_C", "k")
par_ranges <- matrix(c(-4, -2,
                       -4, -2,
                       0, 2,
                       0, 1,
                       0, 1,
                       -2, 0,
                       0, 21),
                     nrow = 2)
colnames(par_ranges) <- raw_par_names
rownames(par_ranges) <- c("min", "max")
N_samples <- 100
par_names <- c("p_C", "p_E", "dstar_C", "dstar_E", "Delta", "two_k_minus_one", "C_on_E")
```

```{r eval = TRUE}
# LHS sampling of transformed parameters from uniform distribution defined by range above
raw_samples <- lhs::randomLHS(N_samples, length(raw_par_names))
colnames(raw_samples) <- raw_par_names
raw_samples[,c("d_E_scaled", "dstar_E_scaled")] <- 
  t(apply(raw_samples[,c("d_E_scaled", "dstar_E_scaled")], 1, sort))

scale_samples <- function(vec, bounds) {
  vec * diff(bounds) + bounds[1]
}

for(i in raw_par_names) {
  raw_samples[,i] <- scale_samples(raw_samples[,i], par_ranges[,i])
}

raw_samples[,"k"] <- ifelse(k0, 0, floor(raw_samples[,"k"]))
samples <- raw_samples
colnames(samples) <- par_names

# untransform to retrieve original parameter values
samples[,"p_C"] <- 10^raw_samples[,"log10_p_C"]
samples[,"p_E"] <- 10^raw_samples[,"log10_p_E"]
samples[,"Delta"] <- 10^raw_samples[,"log10_Delta_on_p_C"] * samples[,"p_C"]
samples[,"dstar_C"] <- 10^raw_samples[,"log10_dstar_C_on_p_C_minus_Delta"] * (samples[,"p_C"] - samples[,"Delta"])
samples[,"dstar_E"] <- 10^(raw_samples[,"log10_p_E"] + (log10(delta) - raw_samples[,"log10_p_E"]) * raw_samples[,"dstar_E_scaled"])
samples[,"two_k_minus_one"] <- 2^raw_samples[,"k"] - 1
d_E <- 10^(raw_samples[,"log10_p_E"] + (log10(delta) - raw_samples[,"log10_p_E"]) * raw_samples[,"d_E_scaled"])
samples[,"C_on_E"] <- (d_E - samples[,"p_E"]) / samples[,"Delta"] / 2^raw_samples[,"k"]
saveRDS(samples, samples_filename)
```

## Solve for fraction of label in downstream compartment

```{r}
samples <- readRDS(samples_filename)
pred_times <- seq(0, 100)
data_list_pars <- lapply(par_names, function(x) array(samples[, x]))
names(data_list_pars) <- par_names

data_list <- c(data_list_pars,
               list(f = f,
                    delta = delta,
                    b_w = b_w,
                    label_end = label_end,
                    N_samples = N_samples,
                    T = length(pred_times),
                    ts = pred_times))
```
```{r eval = TRUE}
pred <- make_pred(pred_model, data_list) %>%
  extract_pred(N_samples = N_samples, pred_name = "Ehat") %>%
  as.data.frame %>% as_tibble %>%
  mutate(t = pred_times) %>%
  pivot_longer(-t, names_to = "run", values_to = "Ehat") %>%
  mutate(run = gsub("V", "", run),
         run = as.numeric(run))
saveRDS(pred, pred_filename)
```

## Add noise to generate simulated data

```{r eval = TRUE}
sim_data <- pred %>%
  mutate(Lhat = vnapply(Ehat, function(y) rlnorm(1, log(y), sigma_L)),
         Lhat = pmax(Lhat, 0))
saveRDS(sim_data, sim_data_filename)
```

## Fit one-compartment model to downstream data

```{r}
sim_data <- sim_data_filename %>% readRDS
fit_single <- function(run1, eval = TRUE) {
  fit_filename <- paste0(dir_name, "fit", run1, ".rds")
  if(!eval) {
    return(fit_filename)
  }
  sim_data <- sim_data %>% filter(run == run1 & t > 0)
  fit_data_list <- c(list(T = nrow(sim_data),
                          ts = sim_data$t,
                          L = sim_data$Lhat,
                          sigma = sigma_L,
                          frac = data_list$f),
                     data_list[c("label_end", "delta")])
  
  model <- "lymphocyte_model_fixed_U_lognormal_noise"
  
  fit <- fit_stan_defaults(model, fit_data_list)
  saveRDS(fit, fit_filename)
  invisible(fit)
}
```

```{r eval = TRUE}
invisible(lapply(seq_len(N_samples), fit_single, eval = TRUE))
```

```{r eval = TRUE}
get_p_median <- function(run1) {
  fit <- run1 %>% fit_single(eval = FALSE) %>% readRDS
  fit <- as.matrix(fit$fit) %>% as.data.frame %>% as_tibble %>%
    select(pb_w, dstar) %>%
    mutate(run = run1,
           p = pb_w / b_w) %>%
    select(-pb_w)
  fit
}
p <- lapply(seq_len(N_samples), get_p_median) %>%
  do.call(rbind, .)
saveRDS(p, p_filename)
```

## Compile file with sampled and fitted values

```{r eval = TRUE}
samples <- readRDS(samples_filename) %>%
  as.data.frame %>% as_tibble %>%
  mutate(run = seq_len(nrow(samples)),
         turnover = (two_k_minus_one + 1) * Delta * C_on_E + p_E,
         production_rate_by_division = two_k_minus_one * Delta * C_on_E + p_E)

get_sum_stats <- function(run1) {
  fit <- run1 %>% fit_single(eval = FALSE) %>% readRDS
  fit <- as.matrix(fit$fit) %>% as.data.frame %>% as_tibble %>%
    select(pb_w, dstar) %>%
    mutate(p = pb_w / b_w) %>%
    select(-pb_w) %>%
    summarise_all(list(lower = function(x) quantile(x, probs = 0.025),
                       median = median,
                       upper = function(x) quantile(x, probs = 0.975))) %>%
    mutate(run = run1)
  fit
}
sum_stats <- lapply(seq_len(N_samples), get_sum_stats) %>%
  do.call(rbind, .) %>%
  full_join(samples) %>%
  rename(p_C = "p_C", p_E = "p_E", dstar_C = "dstar_C", dstar_E = "dstar_E",
         C_on_E = "C_on_E") %>%
  select(run, everything())
write.csv(sum_stats, paste0(dir_name, "source_term_model_estimated_pars.csv"), row.names = FALSE)
```

# Fig 3

```{r}
p <- readRDS(p_filename) %>%
  group_by(run) %>%
  summarise(p_median = median(p), dstar_median = median(dstar)) %>%
  full_join(samples)

if(!k0) {
  p <- p %>% filter(two_k_minus_one > 0)
}

diag_tibble <- tibble(x = c(0, 1))

make_plot <- function(p, diag_tibble, x_var, y_var, x_lab, y_lab, tag_str) {
  x_max <- p %>%
    select(c(x_var, y_var)) %>%
    as.matrix %>%
    max
  ggplot(p, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point() +
    geom_line(data = diag_tibble, aes(x = x, y = x)) +
    theme_bw() +
    coord_cartesian(xlim = c(0, x_max), ylim = c(0, x_max)) +
    theme(aspect.ratio = 1) +
    xlab(x_lab) +
    ylab(y_lab) +
    labs(tag = tag_str)
}

proliferation_plot <- make_plot(p, diag_tibble, "p_E", "p_median",
                                "True value of \n proliferation rate (per day)",
                                latex2exp::TeX("$\\hat{p_E}$ (per day)"),
                                ifelse(k0, "E", "A"))

turnover_plot <- make_plot(p, diag_tibble, "turnover", "p_median",
                           "True value of \n turnover rate (per day)",
                           latex2exp::TeX("$\\hat{p_E}$ (per day)"),
                           ifelse(k0, "F", "B"))

production_plot <- make_plot(p, diag_tibble, "production_rate_by_division", "p_median",
                             "True value of production rate \n by division (per day)",
                             latex2exp::TeX("$\\hat{p_E}$ (per day)"),
                             ifelse(k0, "G", "C"))

dstar_plot <- make_plot(p, diag_tibble, "dstar_E", "dstar_median",
                        "True value of disappearance rate \n of labelled cells (per day)",
                        latex2exp::TeX("$\\hat{d^*_E}$ (per day)"),
                        ifelse(k0, "H", "D"))

g1 <- gridExtra::grid.arrange(proliferation_plot, turnover_plot, production_plot, dstar_plot, nrow = 1)
ggsave(paste0(dir_name, "fit_one_compartment_lhs.pdf"), g1, width = 12, height = 3)
```

```{r}
```
This report is brought to you by `kirlabelling` commit `r system("git -C ~/git_repos/kirlabelling/ rev-parse --short HEAD", intern = TRUE)` and `kirlabbook` commit `r system("git -C ~/2020/kirlabbook/ rev-parse --short HEAD", intern = TRUE)`.