---
title: "Replicate Westera et al. study"
author: "Ada Yan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: TRUE
    keep_tex: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE)
```

# Perfect labelling

## Westera et al. result: in the case of perfect labelling, increasing the number of compartments does not increase the width of the confidence intervals

```{r include = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(deSolve)
# load R package with model fitting functions
devtools::load_all("~/git_repos/cellkineticmodels/") # change this to the directory of the git repository
# working directory should be scripts/ relative to the directory of the git repository
# i.e. the folder this vignette is in

options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)
set.seed(1)
dir_name <- "westera_bayesian_ci_files/"
if(!dir.exists(dir_name)) {
  dir.create(dir_name)
}
```

```{r}
model <- function(t, x, pars) {
  if(t < 0) {
    return(list(0))
  }
  with(as.list(c(x, pars)), {
    U <- as.numeric(t <= label_end)
    dLdt <- p * b_w * U - p * L
    
    return(list(dLdt))
  })
}

solve_label_one_pop <- function(pars) {
  yini <- c(L = 0)
  t1 <- c(seq(0, 150))
  out <- ode(yini, t1, model, pars) %>% as.data.frame %>% as_tibble
  out
}

solve_label_multi_pop <- function(pars) {
  solve_label_inner <- function(pop_idx) {
    pars$p <- pars$p[pop_idx]
    solve_label_one_pop(pars) %>%
      mutate(pop_idx = as.character(pop_idx))
  }
  out <- lapply(seq_along(pars$p), solve_label_inner) %>%
    do.call(rbind, .)
  mean_label <- out %>%
    group_by(time) %>%
    summarise(L = sum(pars[["frac_pop"]] * L)) %>%
    ungroup() %>%
    mutate(pop_idx = "mean")
  out <- bind_rows(out, mean_label)
  out
}
```

```{r}
sim_pars <- list(p = c(0.720, 0.016), frac_pop = c(0.1, 0.9), label_end = 90)

sim_data_wrapper <- function(sim_pars, eval = TRUE) {
  sim_filename <- paste0(dir_name, "sim_data.rds")
  if(!eval) {
    return(sim_filename)
  }
  
  sim_pars <- c(sim_pars, list(b_w = 1, sigma_L = 0.05))
  
  t_vec <- seq_len(21) * 7
  sol <- solve_label_multi_pop(sim_pars)
  
  sim_data <- sol %>%
    filter(pop_idx == "mean" & time %in% t_vec) %>%
    mutate(Lhat = vnapply(L, function(x) rnorm(1, x, sim_pars$sigma_L)))
  saveRDS(list(pars = sim_pars, sol = sol, sim_data = sim_data), sim_filename)
  sim_data
}
```
```{r eval = FALSE}
invisible(sim_data_wrapper(sim_pars, eval = TRUE))
```
```{r}
fit_wrapper <- function(n_compartments = 1, wide_prior, kh = FALSE, eval = TRUE) {
  sim_filename <- sim_data_wrapper(eval = FALSE)
  sim_data <- sim_filename %>%
    readRDS
  fit_filename <- sim_filename %>% gsub("sim_data", "fit", .)
  
  if(wide_prior) {
    fit_filename <- fit_filename %>% gsub("/fit", "/fit_wide_prior", ., fixed = TRUE)
  }
  
  if(n_compartments > 1) {
    fit_filename <- fit_filename %>% gsub("/fit", paste0("/fit_", n_compartments, "_compartments"), ., fixed = TRUE)
  }
  
  if(kh) {
    fit_filename <- fit_filename %>% gsub("/fit", "/fit_kh", ., fixed = TRUE)
  }
  
  if(!eval) {
    return(fit_filename)
  }
  
  sim_data$pars <- c(sim_data$pars, list(perfect_labelling = 1,
                                         frac = .1,
                                         delta = 1,
                                         b_w = 1))
  
  fit_data_list <- c(list(T = nrow(sim_data$sim_data),
                          ts = sim_data$sim_data$time,
                          L = sim_data$sim_data$Lhat,
                          sigma = sim_data$pars$sigma_L,
                          n_pop = n_compartments),
                     sim_data$pars[c("label_end", "delta", "frac", "b_w", "perfect_labelling")])
  
  model <- "lymphocyte_model_fixed_U_n_compartments"
  
  
  if(kh) {
    if(n_compartments > 1) {
      stop("n_compartments must be 1 for kinetic heterogeneity")
    }
    model <- "lymphocyte_model_fixed_U_perfect_labelling"
  } else if(n_compartments == 1) {
    model <- "lymphocyte_model_fixed_U_homogeneous"
  }
  
  seed1 <- 1
  if(wide_prior) {
    model <- paste0(model, "_wide_prior")
  } else if(kh) {
    seed1 <- 2
  }
  
  fit <- fit_stan_defaults(model, fit_data_list, adapt_delta = 0.99, seed = seed1)
  saveRDS(fit, fit_filename)
  invisible(fit)
}
n_compartments_max <- 6
```

```{r eval = FALSE}

lapply(seq_len(n_compartments_max), fit_wrapper, wide_prior = FALSE)
lapply(seq_len(n_compartments_max), fit_wrapper, wide_prior = TRUE)
lapply(c(FALSE, TRUE), fit_wrapper, n_compartments = 1, kh = TRUE)
```

```{r}
plot_p_all <- function(n_compartments_vec, wide_prior) {
  get_p <- function(n_compartments, kh = FALSE) {
    fit <- fit_wrapper(n_compartments = n_compartments, 
                       wide_prior = wide_prior,
                       kh = kh,
                       eval= FALSE) %>%
      readRDS
    sim_data <- sim_data_wrapper(eval = FALSE) %>% readRDS
    
    p <- fit$fit %>% extract_fit(0, c(ifelse(n_compartments == 1 || kh, "p", "bar_p"), "lp__")) %>%
      as.data.frame %>% as_tibble %>%
      mutate(n_compartments = ifelse(kh, "implicit", n_compartments))
    if(n_compartments == 1 || kh) {
      p <- p %>% rename(bar_p = "p")
    }
    
    max_LL_p <- p %>% slice(which.max(lp__)) %>% pull(bar_p)
    median_p <- median(p$bar_p)
    true_p <- sum(sim_data$pars$p * sim_data$pars$frac_pop)
    list(p = p, true_p = true_p)
  }
  
  out <- lapply(n_compartments_vec, get_p)
  out_kh <- get_p(1, TRUE)
  out <- c(out, list(out_kh))
  
  p <- lapply(out, function(x) x$p) %>%
    do.call(rbind, .) %>%
    group_by(n_compartments) %>%
    summarise(lower = quantile(bar_p, 0.025),
              median = median(bar_p),
              upper = quantile(bar_p, 0.975)) %>%
    ungroup()
  true_p <- out[[1]]$true_p
  
  tag1 <- if(wide_prior) "B" else "A"
  ggplot(p, aes(x = n_compartments, group = n_compartments)) +
    geom_point(aes(y = median)) +
    geom_errorbar(aes(ymin = lower, ymax = upper)) +
    geom_hline(yintercept = true_p) +
    theme_bw() +
    expand_limits(y = 0) +
    labs(color = "") +
    xlab("No. of compartments") +
    ylab("Mean proliferation rate (/day)") +
    labs(tag = tag1)
}
```

```{r plot_p_all, fig.width = 5, fig.height = 3, fig.cap = "Posterior distributions for the mean proliferation rate, for different numbers of fitted compartments, with perfect labelling and a prior for p between 0 and 1.  kh denotes fitting a model with kinetic heterogeneity.\\label{fig:plot_p_all}"}
plot_p_all(seq_len(n_compartments_max), wide_prior = FALSE)
```

```{r plot_p_all_wide_prior, fig.width = 5, fig.height = 3, fig.cap = "Posterior distributions for the mean proliferation rate, for different numbers of fitted compartments, with perfect labelling and a prior for p between 0 and 10.  kh denotes fitting a model with kinetic heterogeneity.\\label{fig:plot_p_all_wide_prior}"}
plot_p_all(seq_len(n_compartments_max), wide_prior = TRUE)
```

```{r}
```

This report is brought to you by `kirlabelling` commit `r system("git -C ~/git_repos/kirlabelling/ rev-parse --short HEAD", intern = TRUE)` and `kirlabbook` commit `r system("git -C ~/2020/kirlabbook/ rev-parse --short HEAD", intern = TRUE)`.
